---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
keep_md: true
---


## Loading and preprocessing the data
```{r process}
library(tidyverse)
library(Hmisc)

if(!file.exists('activity.csv')){
        unzip('activity.zip')
}
dat <- read.csv("activity.csv")

dat$date <- as.Date(as.character(dat$date), "%Y-%m-%d") #change date from chr to date format 

```


## What is mean total number of steps taken per day?
```{r totalsteps}
#steps grouped by date and sum calculated
totalsteps <- dat %>% 
        select(steps, date) %>%
        group_by(date) %>%
        summarise(tsteps = sum(steps, na.rm = T))

hist(totalsteps$tsteps, main = "Total Number of Steps Taken Per Day", xlab = "Total number of steps")

meantsteps <- mean(totalsteps$tsteps, na.rm = T)
mediantsteps <- median(totalsteps$tsteps, na.rm = T)
       
```

The mean of the total number of steps taken per day is `r meantsteps` and the medain is `r mediantsteps`.

## What is the average daily activity pattern?
```{r pattern}
#steps grouped by interval and mean for each interval calculated
avgsteps5 <- dat %>%
        select(steps, interval) %>%
        group_by(interval) %>%
        summarise(avgsteps = mean(steps, na.rm = T))

ggplot(avgsteps5, aes(x=interval, y=avgsteps))+
        geom_line(col="blue")+
        labs(x = "Interval", y = "Average number of steps", title = "Number of Steps in Each Five-minute Interval")

int <- avgsteps5[which.max(avgsteps5$avgsteps),] #identify row number with max number of average steps
maxsteps <- int$interval
maxsteps
```
The five-minute interval that contains the maximum number of steps is `r maxsteps` (i.e. 8:35am).

## Imputing missing values
```{r imputing1}

incomplete <- dat[!complete.cases(dat),] #rows with missing values
missing <- nrow(incomplete) #number of rows with missing values
```
The total number of missing values in the dataset (i.e. the total number of rows with NAs) is `r missing`.

```{r imputing2}
imputeddat <- dat
imputeddat$steps <- impute(imputeddat$steps, fun = mean) #missing values for "steps" were filled in with the mean value

#check missing values in other variables
length(which(is.na(imputeddat$steps)))
length(which(is.na(imputeddat$interval)))
length(which(is.na(imputeddat$date)))
#No missing values left

newdat <- imputeddat #new dataset created with missing values filled in

#steps grouped by date and total calculated
totalstepsnew <- newdat %>% 
        select(steps, date) %>%
        group_by(date) %>%
        summarise(tstepsnew = sum(steps))

hist(totalstepsnew$tstepsnew, main = "Total Number of Steps Taken Per Day", xlab = "Total number of steps")

meannew <- mean(totalstepsnew$tstepsnew)
mediannew <- median(totalstepsnew$tstepsnew)

meandiff <- meannew - meantsteps
mediandiff <- mediannew - mediantsteps

```
The mean of total number of steps taken per day is `r meannew` and the median is `r mediannew`.
The new mean and median values are different the estimates that were done before imputing missing data.
The new mean value is `r meandiff` greater and the new median is `r mediandiff` greater than the previous respective estimates.


## Are there differences in activity patterns between weekdays and weekends?
```{r weekdays}
#to create a new variable "days" (as factor) - weekdays or weekends
newdat$days <- ifelse((weekdays(newdat$date) == "Saturday") | (weekdays(newdat$date) =="Sunday"), "weekend", "weekday")
newdat$days <- as.factor(newdat$days)

#steps grouped by interval and days(weekdays/weekends) and average steps calculated
avgstepsdays <- newdat %>%
        select(steps, interval, days) %>%
        group_by(interval,days) %>%
        summarise(avgstepsd = mean(steps, na.rm = T))

ggplot(avgstepsdays, aes(x=interval, y=avgstepsd, col = days))+
        geom_line()+
        facet_grid(cols = vars(days))+
        theme(legend.position="none")+
        labs(x = "Interval", y = "Number of average steps", title = "Difference Activity Patterns between Weekdays and Weekends")